---
title: "Interactions and Heteroskedasticity"
subtitle: "Non-Additivity, WLS, and Robust Inference"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    chalkboard: true
    smaller: true
    scrollable: true
    math: katex
    transition: fade
    width: 1200
    height: 750
    footer: "POL 682 | Interactions & Heteroskedasticity"
---

# Part I: Interactions {background-color="white" style="color: black;"}

What happens when the effect of one predictor depends on the value of another?

## Motivation

Up to this point, we've assumed that the linear regression model is **additive** — the effect of one predictor on $Y$ doesn't depend on the value of another predictor. But this assumption is often unrealistic.

. . .

- Does the effect of authoritarianism on institutional trust differ across partisan groups?
- Does the relationship between income and vote margins change depending on a community's racial composition?

. . .

These are questions about **interactions**. In this lecture, we develop the interaction model, show how it generalizes the additive model, and then turn to what happens when a core assumption of the linear model — constant error variance — is violated.

## Recap: The Additive Model

Recall the additive model with dummy variables from our previous lecture:

$$
Y_i = \alpha + \gamma_{\text{Rep}} D_{\text{Rep}} + \gamma_{\text{Dem}} D_{\text{Dem}} + \beta_{\text{Auth}} X_{\text{Auth}} + e_i
$$

. . .

The expected values for each group:

$$
\begin{eqnarray*}
E(Y \mid \text{Independent}) &=& \alpha + \beta_{\text{Auth}} X_{\text{Auth}} \\
E(Y \mid \text{Republican}) &=& (\alpha + \gamma_{\text{Rep}}) + \beta_{\text{Auth}} X_{\text{Auth}} \\
E(Y \mid \text{Democrat}) &=& (\alpha + \gamma_{\text{Dem}}) + \beta_{\text{Auth}} X_{\text{Auth}}
\end{eqnarray*}
$$

. . .

The slope on authoritarianism is the **same** for every group — only the intercept shifts. The lines are parallel.

## The Problem with Additivity

The additive model forces:

$$\frac{\partial Y}{\partial X_{\text{Auth}}} = \beta_{\text{Auth}}$$

. . .

This is a constant — the same for Republicans, Democrats, and Independents. But what if the effect of authoritarianism *differs* across party groups?

. . .

A Republican who scores high on authoritarianism might have a very different level of institutional trust than a Democrat who scores high on authoritarianism — not just a different *baseline*, but a different *response* to authoritarianism.

. . .

The additive model cannot capture this. We need **interactions**.

## The Interaction Model

We relax the additivity assumption by estimating an **interactive** model:

$$
Y_i = \alpha + \gamma_{\text{Rep}} D_{\text{Rep}} + \gamma_{\text{Dem}} D_{\text{Dem}} + \beta_{\text{Auth}} X_{\text{Auth}} + \delta_{\text{Rep}} (X_{\text{Auth}} \times D_{\text{Rep}}) + \delta_{\text{Dem}} (X_{\text{Auth}} \times D_{\text{Dem}}) + e_i
$$

. . .

We are creating two additional variables: authoritarianism $\times$ Republican and authoritarianism $\times$ Democrat. Just the product of the two variables.

## Interaction Model: Conditional Expectations

$$
\begin{eqnarray*}
E(Y \mid \text{Independent}) &=& \alpha + \beta_{\text{Auth}} X_{\text{Auth}} \\
E(Y \mid \text{Republican}) &=& (\alpha + \gamma_{\text{Rep}}) + (\beta_{\text{Auth}} + \delta_{\text{Rep}}) X_{\text{Auth}} \\
E(Y \mid \text{Democrat}) &=& (\alpha + \gamma_{\text{Dem}}) + (\beta_{\text{Auth}} + \delta_{\text{Dem}}) X_{\text{Auth}}
\end{eqnarray*}
$$

. . .

The marginal effect of authoritarianism now **depends on party identification**:

$$\frac{\partial Y}{\partial X_{\text{Auth}}} = \beta_{\text{Auth}} + \delta_{\text{Rep}} D_{\text{Rep}} + \delta_{\text{Dem}} D_{\text{Dem}}$$

. . .

- $\beta_{\text{Auth}}$ = slope for Independents (reference group)
- $\delta_{\text{Rep}}$ = **difference** in slope between Republicans and Independents
- $\delta_{\text{Dem}}$ = **difference** in slope between Democrats and Independents

The lines are no longer parallel — each group gets its own intercept *and* slope.

## Always Include Lower-Order Terms

It is important to always include the lower-order constituent terms in an interactive model.

. . .

Omitting dummies forces all groups to share the same intercept:

$$Y_i = \alpha + \beta_{\text{Auth}} X_{\text{Auth}} + \delta_{\text{Rep}} (X_{\text{Auth}} \times D_{\text{Rep}}) + \delta_{\text{Dem}} (X_{\text{Auth}} \times D_{\text{Dem}}) + e_i$$

Omitting the main effect of authoritarianism forces the slope to be zero for Independents:

$$Y_i = \alpha + \gamma_{\text{Rep}} D_{\text{Rep}} + \gamma_{\text{Dem}} D_{\text{Dem}} + \delta_{\text{Rep}} (X_{\text{Auth}} \times D_{\text{Rep}}) + \delta_{\text{Dem}} (X_{\text{Auth}} \times D_{\text{Dem}}) + e_i$$

. . .

The full model should include **all constituent terms**. We can always *test* whether they are zero, but we must estimate the full model first.

## Interaction Model: Estimation

```{r, echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE}
library(dplyr)
library(tidyr)

download.file("https://raw.githubusercontent.com/crweber9874/advancedRegression/main/data/wss20.rda",
              destfile = "wss20.rda")
load("wss20.rda")

wss20 <- wss20 |>
  pivot_wider(names_from = contestation, values_from = contestation_value) |>
  mutate(
    authoritarianism = rowMeans(cbind(auth_1, auth_2, auth_3, auth_4), na.rm = TRUE),
    republican = ifelse(party_identification3 == 3, 1, 0),
    democrat = ifelse(party_identification3 == 1, 1, 0),
    independent = ifelse(party_identification3 == 2, 1, 0),
    trust_raw = rowMeans(cbind(trustCongress, trustPresident, trustSC,
                          trustGovernment, trustStateleg, trustPolice,
                          trustScience), na.rm = TRUE),
    institutional_trust = (trust_raw - 1) / 3
  )
wss20$pid_label <- ifelse(wss20$republican == 1 & !is.na(wss20$republican), "Republican",
                   ifelse(wss20$democrat == 1 & !is.na(wss20$democrat), "Democrat",
                   ifelse(wss20$independent == 1 & !is.na(wss20$independent), "Independent", NA)))
```

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# Additive model (restricted)
fit_additive <- lm(institutional_trust ~ authoritarianism + republican + democrat, data = wss20)

# Interactive model (unrestricted)
fit_interaction <- lm(institutional_trust ~ authoritarianism * republican + authoritarianism * democrat, data = wss20)
summary(fit_interaction)
```

## Interaction Model: F-Test

Do the interaction terms jointly improve the model? We can test this with the F-test, comparing the restricted (additive) and unrestricted (interactive) models.

```{r, echo = TRUE, message = FALSE, warning = FALSE}
anova(fit_additive, fit_interaction)
```

. . .

If the F-statistic is significant, the interaction terms improve the model — the slopes genuinely differ across groups.

## Interaction Model: Visualization

```{r, echo = FALSE, message = FALSE, warning = FALSE}
#| fig-width: 10
#| fig-height: 6
library(ggplot2)

b <- coef(fit_interaction)
x_range <- seq(min(wss20$authoritarianism, na.rm = TRUE),
               max(wss20$authoritarianism, na.rm = TRUE), length.out = 100)

line_df <- data.frame(
  pid_label = rep(c("Independent", "Republican", "Democrat"), each = 100),
  x = rep(x_range, 3)
)
line_df$y <- ifelse(line_df$pid_label == "Independent",
               b[1] + b[2] * line_df$x,
             ifelse(line_df$pid_label == "Republican",
               (b[1] + b[3]) + (b[2] + b[5]) * line_df$x,
               (b[1] + b[4]) + (b[2] + b[6]) * line_df$x))

scatter_df <- wss20 |> dplyr::filter(!is.na(pid_label), !is.na(authoritarianism),
                              !is.na(institutional_trust))

ggplot() +
  geom_jitter(data = scatter_df,
              aes(x = authoritarianism, y = institutional_trust, color = pid_label),
              alpha = 0.25, size = 1, width = 0.1, height = 0.1) +
  geom_line(data = line_df, aes(x = x, y = y, color = pid_label), linewidth = 1.2) +
  scale_color_manual(
    values = c("Independent" = "purple", "Republican" = "red", "Democrat" = "blue")
  ) +
  labs(x = "Authoritarianism",
       y = "E(Institutional Trust)",
       color = NULL) +
  theme_minimal() +
  theme(legend.position = c(0.02, y = 0.98), legend.justification = c(0, 1))
```

Unlike the additive model, slopes now differ across groups. Compare this to the parallel lines from the intercept-shift model.

# Continuous-by-Continuous Interactions {background-color="white" style="color: black;"}

What if both variables in the interaction are continuous?

## Setting Up the Arizona Precinct Data

```{r, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE}
load("/Users/chrisweber/Dropbox/github_repos/linear_regression/quarto-book/precinct_voter_summary.rda")
load("/Users/chrisweber/Dropbox/github_repos/linear_regression/quarto-book/precinct_tract_data.rda")

precinct_voter_summary <- precinct_voter_summary |>
  dplyr::mutate(
    pct_latino = (tract_acs_latino / tract_acs_total_population) * 100,
    pct_white  = (tract_acs_non_latino_white / tract_acs_total_population) * 100
  )
head(precinct_voter_summary[, c("dos_precinct_key", "trump_harris_margin",
                                 "tract_acs_median_household_income",
                                 "pct_latino", "tract_acs_median_age",
                                 "tract_acs_gini_index")])
```

## Arizona Precincts: Map

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(sf)
library(leaflet)

az_map <- precinct_tract_data |>
  dplyr::left_join(
    precinct_voter_summary[, c("dos_precinct_key", "trump_harris_margin",
                                "tract_acs_median_household_income",
                                "pct_latino", "tract_acs_median_age",
                                "tract_acs_gini_index")],
    by = "dos_precinct_key"
  ) |>
  dplyr::filter(!is.na(trump_harris_margin))

az_map <- st_transform(az_map, 4326)
az_map <- az_map[st_geometry_type(az_map) %in% c("POLYGON", "MULTIPOLYGON"), ]

pal <- colorNumeric(
  palette = c("blue", "white", "red"),
  domain = az_map$trump_harris_margin
)

leaflet(az_map, width = "100%", height = 500) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addPolygons(
    fillColor = ~pal(trump_harris_margin),
    fillOpacity = 0.7,
    color = "#444", weight = 0.3, opacity = 0.5,
    label = ~paste0(
      "<b>", precinct_name, "</b> (", county, ")<br>",
      "Margin: ", round(trump_harris_margin * 100, 1), " pts<br>",
      "Median Income: $", formatC(tract_acs_median_household_income, format = "d", big.mark = ","), "<br>",
      "% Latino: ", round(pct_latino, 1), "<br>",
      "Median Age: ", round(tract_acs_median_age, 1), "<br>",
      "Gini: ", round(tract_acs_gini_index, 3)
    ) |> lapply(htmltools::HTML),
    highlightOptions = highlightOptions(weight = 2, color = "black", bringToFront = TRUE)
  ) |>
  addLegend("bottomright", pal = pal, values = ~trump_harris_margin,
            title = "Trump\u2013Harris<br>Margin", opacity = 0.7)
```

## Continuous Interaction: The Question

We might ask: does the effect of income on Trump's margin depend on the community's Latino population share?

. . .

In a predominantly white precinct, higher income might push margins in one direction; in a predominantly Latino precinct, the income effect might be different — or even reversed.

. . .

$$
\text{Margin}_i = \beta_0 + \beta_1 \text{Income}_i + \beta_2 \text{Latino}_i + \beta_3 (\text{Income}_i \times \text{Latino}_i) + \beta_4 \text{Age}_i + \beta_5 \text{Gini}_i + e_i
$$

## Continuous Interaction: The Marginal Effect

Take the partial derivative with respect to income:

$$\frac{\partial \text{Margin}}{\partial \text{Income}} = \beta_1 + \beta_3 \times \text{Latino}_i$$

. . .

This is the key insight: the marginal effect of income is *not* a single number. It's a function of percent Latino.

- At a precinct where $\text{Latino} = 0$: effect of income is just $\beta_1$
- At a precinct where $\text{Latino} = 50$: effect is $\beta_1 + 50\beta_3$

. . .

The interaction coefficient $\beta_3$ tells you how much the income effect *changes* for each one-unit increase in percent Latino — but $\beta_3$ by itself doesn't tell you what the income effect actually *is* at any particular value.

. . .

You need both $\beta_1$ and $\beta_3$ together. This is why you can't just look at the coefficient table.

## Continuous Interaction: Estimation

```{r, echo = TRUE, message = FALSE, warning = FALSE}
fit_interact_az <- lm(trump_harris_margin ~ tract_acs_median_household_income *
                        pct_latino + tract_acs_median_age + tract_acs_gini_index,
                      data = precinct_voter_summary)
summary(fit_interact_az)
```

## Computing Marginal Effects

We plug in different values of percent Latino, compute the marginal effect at each one, and plot it. The code below also computes confidence intervals using simulation from the sampling distribution — 1,000 draws from the multivariate normal using `MASS::mvrnorm()`.

```{r, echo = TRUE, message = FALSE, warning = FALSE}
set.seed(42)
b <- coef(fit_interact_az)
v <- vcov(fit_interact_az)
sim_coefs <- MASS::mvrnorm(1000, mu = b, Sigma = v)

# Values of pct_latino to evaluate
latino_vals <- seq(min(precinct_voter_summary$pct_latino, na.rm = TRUE),
                   max(precinct_voter_summary$pct_latino, na.rm = TRUE),
                   length.out = 200)

# For each draw, compute marginal effect: ME = beta_income + beta_interaction * latino
sim_me <- outer(latino_vals, 1:1000, function(lat, i) {
  sim_coefs[i, "tract_acs_median_household_income"] +
    sim_coefs[i, "tract_acs_median_household_income:pct_latino"] * lat
})

me_df <- data.frame(
  pct_latino = latino_vals,
  effect = rowMeans(sim_me),
  lwr = apply(sim_me, 1, quantile, 0.025),
  upr = apply(sim_me, 1, quantile, 0.975)
)
```

## Marginal Effect Plot

```{r, echo = FALSE, message = FALSE, warning = FALSE}
#| fig-width: 10
#| fig-height: 6
library(ggplot2)

ggplot(me_df, aes(x = pct_latino, y = effect)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "steelblue", alpha = 0.2) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Percent Latino",
       y = "Marginal Effect of Median Income\non Trump\u2013Harris Margin",
       title = "How the Income Effect Changes with Latino Population Share") +
  theme_minimal()
```

. . .

The black line is the estimated marginal effect. The shaded band is the 95% CI. Where the band includes zero, the income effect is not statistically distinguishable from zero. Where it's entirely above or below zero, the effect is significant at the 5% level.

## Reading Marginal Effect Plots

This is the right way to interpret a continuous interaction:

. . .

1. The coefficient on the interaction term alone — $\hat{\beta}_3$ — tells you the *rate of change* of the income effect as Latino percentage increases
2. It does **not** tell you what the income effect *is* at any particular value
3. You always need the full marginal effect calculation and the plot

. . .

In practice:

- Compute marginal effects at representative values (e.g., 10th, 25th, 50th, 75th, 90th percentiles)
- Use `vcov()` or simulation to get confidence intervals
- Plot the marginal effect with uncertainty bands

## Interactions: Summary

1. **Categorical interactions** allow different slopes for different groups
2. **Continuous interactions** allow the marginal effect of one variable to vary smoothly with another
3. Always include the lower-order constituent terms
4. Use **marginal effect plots** to interpret continuous interactions — the coefficient on the interaction term alone is not sufficient
5. The **F-test** (`anova()`) tests whether interaction terms jointly improve the model

. . .

Now, let's turn to a different problem. Looking at the Arizona precinct data, you may have noticed something: the *spread* of the residuals isn't constant. That brings us to heteroskedasticity.

# Part II: Heteroskedasticity {background-color="white" style="color: black;"}

When the error variance is not constant

## Recall the Gauss-Markov Assumptions

| # | Assumption | Math | Purpose |
|:--:|:-----------|:----|:--------|
| 1 | Linearity | $y_i = \mathbf{x}_i^T\boldsymbol{\beta} + \epsilon_i$ | Estimation |
| 2 | Exogenous IVs | $\text{cov}(X_i, \epsilon_i) = 0$ | Estimation |
| 3 | Zero mean error | $E(\epsilon_i) = 0$ | Inference |
| **4** | **Homoskedasticity** | $\text{var}(\epsilon_i) = \sigma^2$ | **Inference** |
| 5 | Normality | $\epsilon_i \sim N(0, \sigma)$ | Inference |
| 6 | Independent errors | $\text{cov}(\epsilon_i, \epsilon_j) = 0$ | Inference |
| 7 | Variation in X | $\text{var}(X) > 0$ | Estimation |
| 8 | Correct specification | No omitted/extraneous variables | Estimation |
| 9 | No perfect multicollinearity | No perfect linear combinations | Estimation |

. . .

Assumptions 1–2 and 7–9 are needed for **unbiased** estimates. Assumptions 3–6 are needed for valid **inference**. Today we focus on Assumption 4.

## The Variance-Covariance Matrix under Homoskedasticity

Under the Gauss-Markov assumptions:

$$E(\boldsymbol{\epsilon}\boldsymbol{\epsilon}^T) = \sigma^2 \mathbf{I} = \begin{bmatrix} \sigma^2&\cdots & 0\\ 0&\sigma^2 & 0\\ \vdots & \vdots & \vdots\\ 0&\cdots &\sigma^2\\ \end{bmatrix}$$

. . .

Every observation has the same error variance, and errors are uncorrelated. The variance-covariance matrix of the OLS estimator simplifies neatly:

$$\text{var}(\mathbf{b}) = \sigma^2 (\mathbf{X}^T\mathbf{X})^{-1}$$

This is what R reports when you call `summary(lm(...))`.

## What Goes Wrong with Heteroskedasticity?

If $E(\boldsymbol{\epsilon}\boldsymbol{\epsilon}^T) = \sigma^2 \boldsymbol{\Omega}$ where $\boldsymbol{\Omega} \neq \mathbf{I}$, then:

$$\text{var}(\mathbf{b}) = \sigma^2(\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\boldsymbol{\Omega} \mathbf{X}(\mathbf{X}^T\mathbf{X})^{-1}$$

. . .

The $\boldsymbol{\Omega}$ sits in the middle and **does not simplify**. The standard formula $\sigma^2 (\mathbf{X}^T\mathbf{X})^{-1}$ is **wrong**.

. . .

**Consequences:**

- OLS coefficients remain **unbiased** — but they are no longer efficient (not BLUE)
- Standard errors are **wrong**
- $t$-values, $p$-values, and confidence intervals are **wrong**
- Type II errors increase — we may fail to reject false nulls

## Sources of Heteroskedasticity

**Heterogeneous units.** Consider a model predicting political contributions from income. Low-income individuals cluster near zero, but high-income individuals vary enormously. The variance fans out like a megaphone.

. . .

**Outliers.** A handful of extreme observations inflate the residual variance in certain regions of the predictor space while leaving it small elsewhere.

. . .

**Incorrect specification.** If the true relationship is nonlinear but we fit a linear model, residuals are systematically larger where the misspecification is worst.

. . .

**Learning effects.** In panel or repeated-measures designs, respondents may become more consistent over time. Early waves show high variability; later waves converge.

# Building the WLS Estimator {background-color="white" style="color: black;"}

Transforming the data to restore constant variance

## The Core Idea

The heteroskedastic error covariance is:

$$E(\boldsymbol{\epsilon}\boldsymbol{\epsilon}^T) = \sigma^2 \boldsymbol{\Omega}$$

where $\boldsymbol{\Omega}$ is a diagonal matrix with unequal elements — some observations are "noisier" than others.

. . .

**The WLS strategy:** transform the data so that the transformed errors have constant variance. Then apply OLS to the transformed data. The transformed OLS estimator will be BLUE.

. . .

If observation $i$ has a large error variance (noisy), we **downweight** it. If observation $j$ has a small error variance (precise), we **upweight** it.

## Step 1: The Weight Matrix

We need $\boldsymbol{\Omega}$, the matrix describing how error variance differs across observations:

$$\boldsymbol{\Omega}=\begin{bmatrix}w_{11}&\cdots & 0\\0&w_{22}& 0\\\vdots & \vdots & \vdots\\0&\cdots &w_{nn}\\\end{bmatrix}$$

. . .

The diagonal entries $w_{ii}$ represent the relative variance of each observation. For instance, a precinct with $w_{ii} = 4$ has four times the error variance of the baseline.

## Step 2: Decompose the Inverse

Define $\boldsymbol{\rho}$ such that $\boldsymbol{\rho}^T\boldsymbol{\rho} = \boldsymbol{\Omega}^{-1}$.

. . .

In scalar algebra, if the variance of $\epsilon_i$ is $\sigma^2 \omega_{ii}$, we'd just divide the entire equation by $\sqrt{\omega_{ii}}$. In matrix form, we premultiply by $\boldsymbol{\rho}$:

$$\rho_{ii} = \frac{1}{\sqrt{\omega_{ii}}}$$

. . .

**Example:** Three observations with $\omega = [1, 4, 9]$:

$$\boldsymbol{\Omega} = \begin{bmatrix} 1 & 0 & 0 \\ 0 & 4 & 0 \\ 0 & 0 & 9 \end{bmatrix}, \quad \boldsymbol{\rho} = \begin{bmatrix} 1 & 0 & 0 \\ 0 & 1/2 & 0 \\ 0 & 0 & 1/3 \end{bmatrix}$$

- Observation 1 ($\omega = 1$): multiplied by 1 — no change (precise)
- Observation 2 ($\omega = 4$): multiplied by $1/2$ — halved (moderate)
- Observation 3 ($\omega = 9$): multiplied by $1/3$ — shrunk most (noisy)

## Step 3: Transform and Estimate

Premultiply the regression equation by $\boldsymbol{\rho}$:

$$\boldsymbol{\rho}\mathbf{y} = \boldsymbol{\rho}\mathbf{X}\boldsymbol{\beta} + \boldsymbol{\rho}\boldsymbol{\epsilon}$$

. . .

Define $\mathbf{y}^* = \boldsymbol{\rho}\mathbf{y}$, $\mathbf{X}^* = \boldsymbol{\rho}\mathbf{X}$, $\boldsymbol{\epsilon}^* = \boldsymbol{\rho}\boldsymbol{\epsilon}$.

. . .

The variance of the transformed error:

$$\text{var}(\epsilon_i^*) = \text{var}\left(\frac{\epsilon_i}{\sqrt{w_{ii}}}\right) = \frac{1}{w_{ii}} \cdot \sigma^2 w_{ii} = \sigma^2$$

. . .

The heteroskedasticity is gone. Every transformed error has the same variance $\sigma^2$.

## The WLS Estimator

Apply OLS to the transformed equation:

$$\mathbf{b}^{*} = (\mathbf{X}^{*T}\mathbf{X}^{*})^{-1}\mathbf{X}^{*T}\mathbf{y}^{*}$$

. . .

Substituting $\mathbf{X}^* = \boldsymbol{\rho}\mathbf{X}$ and $\mathbf{y}^* = \boldsymbol{\rho}\mathbf{y}$, and since $\boldsymbol{\rho}^T\boldsymbol{\rho} = \boldsymbol{\Omega}^{-1}$:

$$\mathbf{b}^{*} = (\mathbf{X}^{T} \boldsymbol{\Omega}^{-1}\mathbf{X})^{-1}\mathbf{X}^{T}\boldsymbol{\Omega}^{-1}\mathbf{y}$$

. . .

This is BLUE: $E(\mathbf{b}^*) = \boldsymbol{\beta}$ with minimum variance among linear unbiased estimators.

The variance of the WLS estimator is:

$$\text{var}(\mathbf{b}^*) = \sigma^2(\mathbf{X}^T\boldsymbol{\Omega}^{-1}\mathbf{X})^{-1}$$

## The Catch: Where Do Weights Come From?

To construct the weights, we need to **know what's causing the heteroskedasticity**. The procedure depends on $\boldsymbol{\Omega}$, but $\boldsymbol{\Omega}$ isn't given to us.

. . .

- If we get $\boldsymbol{\Omega}$ right, WLS is BLUE — the best we can do
- If we get it wrong, WLS may actually perform **worse** than OLS

. . .

Common weighting schemes:

- **Frequency weights**: row represents multiple units; weight = count
- **Probability weights**: survey sampling; $w_i = 1/\pi_i$
- **Aggregated data**: group means from $n_i$ individuals; weight = $n_i$
- **Inverse-variance**: meta-analysis; $w_i = 1/\text{SE}_i^2$
- **Propensity scores**: causal inference; balances treatment/control

# White's Robust Standard Errors {background-color="white" style="color: black;"}

An alternative that doesn't require knowing $\boldsymbol{\Omega}$

## The Idea Behind Robust SEs

Rather than transform the data, estimate the actual error covariance directly. White (1980) shows we can estimate $\boldsymbol{\Sigma}$:

$$\hat{\boldsymbol{\Sigma}} = \begin{bmatrix}e_1^2&\cdots & 0\\0&e_2^2& 0\\\vdots & \vdots & \vdots\\0&\cdots &e_n^2\\\end{bmatrix}$$

. . .

Then plug this into the sandwich formula:

$$(\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T \hat{\boldsymbol{\Sigma}} \mathbf{X}(\mathbf{X}^{T}\mathbf{X})^{-1}$$

. . .

**Advantages:** No a priori knowledge of $\boldsymbol{\Omega}$ needed. Conservative and defensive.

**Disadvantages:** Requires large samples. Less efficient than correctly specified WLS.

## Trade-Offs: WLS vs. Robust SEs

| | WLS | Robust SEs |
|--|-----|-----------|
| **Requires** | Knowledge of $\boldsymbol{\Omega}$ | Nothing beyond OLS |
| **If correct** | BLUE — most efficient | Valid but less efficient |
| **If wrong** | Can be *worse* than OLS | Still valid |
| **Changes** | Coefficients and SEs | Only SEs |
| **Best for** | Known weights (surveys, meta-analysis) | Default defensive strategy |

. . .

It is recommended to attempt multiple approaches and examine if they converge on the same solution.

# Detection {background-color="white" style="color: black;"}

How do we know if heteroskedasticity is present?

## Visual Diagnostics

The first step should always be visual. Plot the fitted values against the residuals. Under homoskedasticity, this should look like a random cloud with constant spread.

. . .

Any systematic pattern — a funnel, a megaphone, a triangle shape — suggests heteroskedasticity.

. . .

We can also plot squared residuals against each predictor. Systematic trends indicate which variable is driving the non-constant variance.

## Goldfeld-Quandt Test

1. Construct the null hypothesis: *no heteroskedasticity*
2. Sort the data in ascending order of the suspected variable
3. Remove $C$ middle observations; create two datasets (low and high)
4. Run two regressions; compare $\text{RSS}_2/\text{RSS}_1 \sim F$

. . .

**Limitation:** Must know the suspected source. Low power in small samples.

## Breusch-Pagan Test

Regress the squared residuals on the independent variables:

$$U_i = \beta_1 + \beta_2 X_{2i} + \beta_3 X_{3i} + \omega_i$$

. . .

**Null:** $H_0: \beta_2 = \beta_3 = \cdots = 0$ (no heteroskedasticity)

**Test statistic:** $\text{RegSS}/2 \sim \chi^2_K$

. . .

If $\chi^2_{\text{computed}} > \chi^2_{\text{critical}}$, reject the null.

## Cook-Weisberg Test

Nearly identical to BP, but regress $U_i$ on the **fitted values** only:

$$U_i = \beta_1 + \beta_2 \hat{Y}_i + \omega_i$$

. . .

- One degree of freedom, so somewhat more **powerful**
- Doesn't tell us *which* variable causes heteroskedasticity
- Still use $\text{RegSS}/2 \sim \chi^2_1$

## White's Test

The most general test. Regress $e_i^2$ on all predictors, their squares, and their cross-products:

$$e_i^2 = \beta_1 + \beta_2 X_2 + \beta_3 X_3 + \beta_4 X_2^2 + \beta_5 X_3^2 + \beta_6 X_2 X_3 + \omega_i$$

. . .

Number of parameters: $P = [K(K+3)/2] + 1$

Test: $n \times R^2 \sim \chi^2_P$

. . .

**Advantage:** No assumption about which variable causes heteroskedasticity.

**Disadvantage:** Requires many parameters; can have low power with many predictors.

# Applied Analysis: Arizona Precinct Data {background-color="white" style="color: black;"}

Putting everything together

## Estimation and Rescaling

We rescale each predictor to a 0–1 range and mean-center so that coefficients are directly comparable.

```{r, echo = TRUE, message = FALSE, warning = FALSE}
rescale01 <- function(x) {
  r <- (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
  r - mean(r, na.rm = TRUE)
}

precinct_voter_summary <- precinct_voter_summary |>
  dplyr::mutate(
    income_sc = rescale01(tract_acs_median_household_income),
    latino_sc = rescale01(pct_latino),
    age_sc    = rescale01(tract_acs_median_age),
    gini_sc   = rescale01(tract_acs_gini_index)
  )

fit_az <- lm(trump_harris_margin ~ income_sc + latino_sc + age_sc + gini_sc,
             data = precinct_voter_summary)
summary(fit_az)
```

## Visual Diagnostics: Residuals vs. Fitted

```{r, echo = TRUE, message = FALSE, warning = FALSE}
#| fig-width: 10
#| fig-height: 5
library(ggplot2)

plot_df <- data.frame(
  fitted = fitted(fit_az),
  residuals = residuals(fit_az)
)

ggplot(plot_df, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.3, size = 1.5, color = "gray40") +
  geom_hline(yintercept = 0, color = "red", linewidth = 0.8) +
  geom_smooth(method = "loess", se = FALSE, color = "steelblue", linewidth = 1) +
  labs(x = expression(hat(Y) ~ "(Fitted Values)"),
       y = "Residuals",
       title = "Residuals vs. Fitted Values") +
  theme_minimal()
```

Notice the spread of residuals changes across the range of fitted values. This is heteroskedasticity.

## Visual Diagnostics: Squared Residuals by Predictor

```{r, message = FALSE, warning = FALSE}
#| echo: true
#| code-fold: true
#| fig-width: 10
#| fig-height: 8
model_data <- fit_az$model
model_data$resid_sq <- residuals(fit_az)^2

p1 <- ggplot(model_data, aes(x = income_sc, y = resid_sq)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "loess", se = FALSE, color = "steelblue") +
  labs(x = "Median Household Income (scaled)", y = expression(e[i]^2)) +
  theme_minimal()

p2 <- ggplot(model_data, aes(x = latino_sc, y = resid_sq)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "loess", se = FALSE, color = "steelblue") +
  labs(x = "Latino Population (scaled)", y = expression(e[i]^2)) +
  theme_minimal()

p3 <- ggplot(model_data, aes(x = age_sc, y = resid_sq)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "loess", se = FALSE, color = "steelblue") +
  labs(x = "Median Age (scaled)", y = expression(e[i]^2)) +
  theme_minimal()

p4 <- ggplot(model_data, aes(x = gini_sc, y = resid_sq)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "loess", se = FALSE, color = "steelblue") +
  labs(x = "Gini Index (scaled)", y = expression(e[i]^2)) +
  theme_minimal()

gridExtra::grid.arrange(p1, p2, p3, p4, ncol = 2)
```

## Formal Tests

```{r, echo = TRUE, message = FALSE, warning = FALSE}
library(lmtest)
library(car)

# Breusch-Pagan Test
cat("=== Breusch-Pagan Test ===\n")
bptest(fit_az)
```

. . .

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# Cook-Weisberg (against fitted values)
cat("=== Cook-Weisberg Test ===\n")
ncvTest(fit_az)
```

## White's Test

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# White's Test (BP against all predictors, squares, and cross-products)
aux_data <- fit_az$model
aux_data$yhat <- fitted(fit_az)
aux_data$inc2  <- aux_data$income_sc^2
aux_data$lat2  <- aux_data$latino_sc^2
aux_data$age2  <- aux_data$age_sc^2
aux_data$gini2 <- aux_data$gini_sc^2

cat("=== White's Test ===\n")
bptest(fit_az, ~ income_sc + latino_sc + age_sc + gini_sc +
                  inc2 + lat2 + age2 + gini2 +
                  income_sc:latino_sc +
                  income_sc:age_sc +
                  income_sc:gini_sc +
                  latino_sc:age_sc +
                  latino_sc:gini_sc +
                  age_sc:gini_sc,
       data = aux_data)
```

. . .

All tests reject the null of constant variance. There is strong evidence of heteroskedasticity in the Arizona precinct data.

## Testing Against Specific Predictors

```{r, echo = TRUE, message = FALSE, warning = FALSE}
cat("=== ncvTest against income ===\n")
ncvTest(fit_az, ~ income_sc)

cat("\n=== ncvTest against latino ===\n")
ncvTest(fit_az, ~ latino_sc)
```

. . .

This helps identify *which* variable drives the heteroskedasticity.

## Residual Plots: `car` Package

```{r, echo = TRUE, message = FALSE, warning = FALSE}
#| fig-width: 10
#| fig-height: 8
residualPlots(fit_az, tests = TRUE)
```

## Robust Standard Errors

```{r, echo = TRUE, message = FALSE, warning = FALSE}
library(sandwich)

cat("=== OLS Standard Errors ===\n")
coeftest(fit_az)
```

## Robust Standard Errors (Continued)

```{r, echo = TRUE, message = FALSE, warning = FALSE}
cat("=== HC1 (White) Robust Standard Errors ===\n")
coeftest(fit_az, vcov = vcovHC(fit_az, type = "HC1"))
```

. . .

The coefficient *estimates* don't change — only the standard errors and therefore the $t$-statistics and $p$-values. If the robust SEs differ substantially from OLS, that's evidence of heteroskedasticity affecting inference.

## WLS Correction: Two-Step Procedure

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# Step 1: Estimate the variance function
resid_sq <- residuals(fit_az)^2
resid_sq[resid_sq == 0] <- min(resid_sq[resid_sq > 0])  # guard against log(0)
wls_data <- fit_az$model
wls_data$log_resid_sq <- log(resid_sq)

var_model <- lm(log_resid_sq ~ income_sc + latino_sc + age_sc + gini_sc,
                data = wls_data)

# Step 2: Compute weights
# sigma_hat_i = sqrt(exp(fitted values))
# weight = 1 / sigma_hat -- matching rho_ii = 1/sqrt(omega_ii)
wls_data$sigma_hat <- sqrt(exp(fitted(var_model)))
wls_data$w <- 1 / wls_data$sigma_hat

# Step 3: WLS model
fit_wls <- glm(trump_harris_margin ~ income_sc + latino_sc + age_sc + gini_sc,
               data = wls_data,
               weights = w,
               family = gaussian())
summary(fit_wls)
```

## Comparing Standard Errors: OLS vs. Robust vs. WLS

```{r, echo = TRUE, message = FALSE, warning = FALSE}
ols_se <- sqrt(diag(vcov(fit_az)))
robust_se <- sqrt(diag(vcovHC(fit_az, type = "HC1")))
wls_se <- sqrt(diag(vcov(fit_wls)))

comparison <- data.frame(
  OLS = round(ols_se, 6),
  Robust_HC1 = round(robust_se, 6),
  WLS = round(wls_se, 6)
)
comparison |> knitr::kable()
```

. . .

Where the robust and WLS standard errors agree with each other but differ from OLS, you can be fairly confident the OLS standard errors are misleading.

## Residual Map

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(sf)
library(leaflet)

az_resid <- data.frame(
  dos_precinct_key = precinct_voter_summary$dos_precinct_key[as.integer(rownames(fit_az$model))],
  resid_sq = residuals(fit_az)^2,
  margin = fit_az$model$trump_harris_margin
)

az_map <- precinct_tract_data |>
  dplyr::left_join(az_resid, by = "dos_precinct_key") |>
  dplyr::filter(!is.na(resid_sq))

az_map <- st_transform(az_map, 4326)
az_map <- az_map[st_geometry_type(az_map) %in% c("POLYGON", "MULTIPOLYGON"), ]

pal <- colorNumeric(palette = "YlOrRd", domain = az_map$resid_sq)

leaflet(az_map, width = "100%", height = 500) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addPolygons(
    fillColor = ~pal(resid_sq),
    fillOpacity = 0.7,
    color = "#444", weight = 0.3, opacity = 0.5,
    label = ~paste0(
      "<b>", precinct_name, "</b> (", county, ")<br>",
      "Margin: ", round(margin * 100, 1), " pts<br>",
      "Squared Residual: ", round(resid_sq, 4)
    ) |> lapply(htmltools::HTML),
    highlightOptions = highlightOptions(weight = 2, color = "black", bringToFront = TRUE)
  ) |>
  addLegend("bottomright", pal = pal, values = ~resid_sq,
            title = "Squared<br>Residual", opacity = 0.7)
```

If heteroskedasticity has a spatial pattern, this can help diagnose the source.

## The Diagnostic Workflow

1. **Fit the model** and examine residual plots for visual signs
2. **Run formal tests** (BP, CW, White) to confirm and identify drivers
3. **Apply corrections** — robust SEs or WLS — and compare results to OLS
4. If robust and WLS agree but differ from OLS, the OLS SEs are misleading

. . .

In practice, robust SEs are the **default defensive strategy**. Use WLS when the variance structure is known or well-motivated.

## Summary

1. **Interactions** relax the additivity assumption — the effect of one predictor depends on the value of another
2. Categorical interactions produce different slopes for different groups; continuous interactions produce marginal effects that vary smoothly
3. Always include lower-order constituent terms; use marginal effect plots for interpretation
4. **Heteroskedasticity** violates Gauss-Markov Assumption 4 — OLS remains unbiased but SEs are wrong
5. **WLS** transforms the data to restore constant variance, producing a BLUE estimator
6. **Robust standard errors** (White, 1980) correct inference without requiring knowledge of $\boldsymbol{\Omega}$
7. Visual diagnostics first, formal tests second, corrections third — always compare approaches

## References

- Breusch, T. S., and A. R. Pagan. 1979. "A Simple Test for Heteroskedasticity and Random Coefficient Variation." *Econometrica* 47(5): 1287–1294.
- Cook, R. D., and S. Weisberg. 1983. "Diagnostics for Heteroscedasticity in Regression." *Biometrika* 70(1): 1–10.
- Fox, John. 2008. *Applied Regression Analysis and Generalized Linear Models*. Sage.
- Fox, John, and Sanford Weisberg. 2019. *An R Companion to Applied Regression*. 3rd ed. Sage.
- Goldfeld, S. M., and R. E. Quandt. 1965. "Some Tests for Homoscedasticity." *Journal of the American Statistical Association* 60(310): 539–547.
- White, Halbert. 1980. "A Heteroskedasticity-Consistent Covariance Matrix Estimator and a Direct Test for Heteroskedasticity." *Econometrica* 48(4): 817–838.
