---
title: "Multiple Regression"
subtitle: "Multiple Predictors and Matrix Algebra"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    chalkboard: true
    smaller: true
    scrollable: true
    math: katex
    transition: fade
    width: 1200
    height: 750
---

## Multiple variables

The OLS estimator is a point estimator. We predict a single point $\hat{y}_i$, given a predictor $x_i$. Let's consider the SRF with *multiple predictors*:

$$
Y_i - \underbrace{(a + b_1 X_1 + b_2 X_2)}_{\hat{Y}_{i}} = e_i
$$

We'll still find a line that minimizes $\sum e^2$


- We can just use a little algebra to rewrite these equations. Let's simplify things by writing each term in "deviation" form.

 * $y_i=Y_i-\bar{Y}$
 * $x_{1i}=X_{1i}-\bar{X_1}$
 * $x_{2i}=X_{2i}-\bar{X_2}$

## Multiple variables

Setting $\frac{\partial SSR}  {\partial b_1}$ to 0

\begin{eqnarray*}
0&=& -2 \sum (Y_i-a-b_1X_{1i}-b_2 X_{2i}) (X_{1i})     \nonumber \\
b_1&=&  \frac{\sum x_{1i} y_i-b_2\sum x_{1i} x_{2i}}{\sum x_{1i}^2} \nonumber \\
\end{eqnarray*}

 Setting $\frac{\partial SSR}  {\partial b_2}$ to 0
\begin{eqnarray*}
0&=& -2 \sum (Y_i-a-b_1X_{1i}-b_2 X_{2i}) (X_{2i})     \nonumber \\
b_2&=&  \frac{\sum x_{2i} y_i-b_1\sum x_{1i} x_{2i}}{\sum x_{2i}^2} \nonumber \\
\end{eqnarray*}

## Multiple variables

Then,

\begin{eqnarray*}
b_1&=&  \frac{\sum y_i x_{1i} \sum x_{2i}^2 - \sum x_{1i} x_{2i} \sum x_{2i} y_i  }{\sum x_{1i}^2 \sum x_{2i}^2 -(\sum x_{1i} x_{2i})^2} \nonumber \\
b_2&=&  \frac{\sum y_i x_{2i} \sum x_{1i}^2 - \sum x_{1i} x_{2i} \sum x_{1i} y_i }{\sum x_{1i}^2 \sum x_{2i}^2 -(\sum x_{1i} x_{2i})^2}
\nonumber \\
\end{eqnarray*}

## Multiple variables

```{r}
#| echo: false
#| output: false
library(plotly)
set.seed(42)

n <- 500
x1 <- rnorm(n, 3, 2)
x2 <- rnorm(n, 1, 2)
y <- 1 + 0.75 * x1 - 0.50 * x2 + rnorm(n, 0, 1.5)

dat <- data.frame(x1, x2, y)
fit <- lm(y ~ x1 + x2, data = dat)
dat$yhat <- fitted(fit)
```

$\hat{Y} = a + b_1 X_1 + b_2 X_2$

- $a$ = `r round(coef(fit)[1], 3)`
- $b_1$ = `r round(coef(fit)[2], 3)`
- $b_2$ = `r round(coef(fit)[3], 3)`

$R^2$ = `r round(summary(fit)$r.squared, 3)`

## Multiple Regression: Fitted Plane & Residuals

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7
# Build plane surface
grd_x1 <- seq(min(x1), max(x1), length.out = 25)
grd_x2 <- seq(min(x2), max(x2), length.out = 25)
grd <- expand.grid(x1 = grd_x1, x2 = grd_x2)
grd$y <- predict(fit, newdata = grd)
z_mat <- matrix(grd$y, nrow = 25, ncol = 25)

# Residual segments: vertical lines from each point to the plane
seg <- do.call(rbind, lapply(1:n, function(i) {
  data.frame(
    sx1 = c(dat$x1[i], dat$x1[i], NA),
    sx2 = c(dat$x2[i], dat$x2[i], NA),
    sz  = c(dat$y[i],  dat$yhat[i], NA)
  )
}))

p <- plot_ly()
p <- add_surface(p, x = grd_x2, y = grd_x1, z = z_mat,
                 colorscale = list(c(0, "#2c7fb8"), c(1, "blue")),
                 opacity = 0.35, showscale = FALSE, name = "Regression Plane")
p <- add_markers(p, x = dat$x2, y = dat$x1, z = dat$y,
                 marker = list(size = 3, color = "black", opacity = 0.6),
                 name = "Observations")
p <- add_trace(p, x = seg$sx2, y = seg$sx1, z = seg$sz,
               type = "scatter3d", mode = "lines",
               line = list(color = "#d62728", width = 2),
               opacity = 0.4, name = "Residuals")
p <- layout(p,
  scene = list(
    xaxis = list(title = "X2"),
    yaxis = list(title = "X1"),
    zaxis = list(title = "Y"),
    camera = list(eye = list(x = 1.8, y = -1.5, z = 0.8))
  ),
  margin = list(l = 0, r = 0, b = 0, t = 0),
  title = ""
)

# Auto-rotate the 3D plot
p <- htmlwidgets::onRender(p, "
  function(el, x) {
    var angle = 0;
    var radius = 2.2;
    var zHeight = 0.8;
    function rotate() {
      angle += 0.002;
      var eye = {
        x: radius * Math.cos(angle),
        y: radius * Math.sin(angle),
        z: zHeight
      };
      Plotly.relayout(el.id, {'scene.camera.eye': eye});
      requestAnimationFrame(rotate);
    }
    rotate();
  }
")
p
```

# Matrix Algebra {background-color="white" style="color: black;"}

Vectors, Matrices, and the OLS Estimator

## Why Matrix Algebra?
- Writing the SRF in matrix form makes it easier to solve
- Quantitative social science aims to quantify relationships between **multiple variables**
- Data is *tabular*: rows = observations, columns = variables
- We need tools to solve systems of equations efficiently

$$\begin{bmatrix}
y_{1}= & b_0+ b_1 x_{1}+b_2 x_{2}\\
y_{2}= & b_0+ b_1 x_{1}+b_2 x_{2}\\
\vdots\\
y_{n}= & b_0+ b_1 x_{1}+b_2 x_{2}
\end{bmatrix}$$

- $n$ equations, fewer unknowns — **linear algebra** gives us the solution

## Data as a Matrix

Each row is an observation; each column is a variable.

$$\begin{bmatrix}
Vote & PID & Ideology \\\hline
a_{11} & a_{12} & a_{13} \\
a_{21} & a_{22} & a_{23} \\
\vdots & \vdots & \vdots\\
a_{n1} & a_{n2} & a_{n3}
\end{bmatrix}$$

- This is an $n \times 3$ matrix
- First subscript = row, second = column
- Notation: $\mathbf{A}_{n \times 3}$

```{r, echo = TRUE}
A <- matrix(c(1, 0, 3,
              0, 1, 5,
              1, 1, 4), nrow = 3, byrow = TRUE)
colnames(A) <- c("Vote", "PID", "Ideology")
A
dim(A)
```

## Vectors: The Building Blocks

- **Scalar**: a single number (magnitude only)
- **Vector**: multiple elements — encodes **magnitude** and **direction**

A vector $\mathbf{a} \in \mathbb{R}^k$ has $k$ elements.

. . .

**Euclidean Distance** between $\mathbf{a}=[x_1,y_1]$ and $\mathbf{b}=[x_2,y_2]$:

$$\text{Distance}(\mathbf{a},\mathbf{b}) = \sqrt{(x_1-x_2)^2+(y_1-y_2)^2}$$

This is just the Pythagorean theorem!

```{r, echo = TRUE}
a <- c(3, 2); b <- c(1, 1)
sqrt(sum((a - b)^2))  # Euclidean distance
```

## The Norm of a Vector

The **norm** measures the length (magnitude) of a vector from the origin:

$$\|\mathbf{a}\| = \sqrt{x_1^2 + y_1^2}$$

- Dividing a vector by its norm gives a **unit vector** (length = 1)
- Useful for standardization

```{r, echo = TRUE}
a <- c(3, 2, 1)
sqrt(sum(a^2))       # norm of a
a / sqrt(sum(a^2))   # unit vector
```

. . .

**In higher dimensions** ($\mathbb{R}^3$):

$$\|\mathbf{a}\| = \sqrt{x_1^2 + y_1^2 + z_1^2}$$

## Vector Addition & Subtraction

Element-wise operations on **conformable** vectors (same length):

$$\mathbf{a} + \mathbf{b} = [3+1,\; 2+1,\; 1+1] = [4, 3, 2]$$

$$\mathbf{a} - \mathbf{b} = [3-1,\; 2-1,\; 1-1] = [2, 1, 0]$$

. . .

**Properties:**

| Property | Statement |
|----------|-----------|
| Commutative | $\mathbf{a}+\mathbf{b}=\mathbf{b}+\mathbf{a}$ |
| Associative | $(\mathbf{a}+\mathbf{b})+\mathbf{c}=\mathbf{a}+(\mathbf{b}+\mathbf{c})$ |
| Distributive | $c(\mathbf{a}+\mathbf{b})=c\mathbf{a}+c\mathbf{b}$ |
| Zero | $\mathbf{a}+0=\mathbf{a}$ |

```{r, echo = TRUE}
a <- c(3, 2, 1); b <- c(1, 1, 1)
a + b    # addition
a - b    # subtraction
```

## Vector Multiplication

- **Inner (dot) product** → produces a *scalar* (measures similarity / covariance)
- **Cross product** → produces a *vector* (orthogonal to both inputs)
- **Outer product** → produces a *matrix*

## The Inner (Dot) Product

Multiply corresponding elements and sum:

$$\mathbf{a} \cdot \mathbf{b} = \sum_i a_i b_i$$

For $\mathbf{a}=[3,2,1]$ and $\mathbf{b}=[1,1,1]$: $\;\; 3(1)+2(1)+1(1) = 6$

```{r, echo = TRUE}
a <- c(3, 2, 1); b <- c(1, 1, 1)
sum(a * b)          # inner product
a %*% b             # same thing with matrix notation
```

. . .

**The inner product is a measure of covariance:**

$$\text{cov}(x,y) = \frac{\text{inner product}(x-\bar{x},\; y-\bar{y})}{n-1}$$

$$r_{x,y} = \frac{\text{inner product}(x-\bar{x},\; y-\bar{y})}{\|x-\bar{x}\|\;\|y-\bar{y}\|}$$


## Inner Product Rules

| Property | Statement |
|----------|-----------|
| Commutative | $\mathbf{a} \cdot \mathbf{b} = \mathbf{b} \cdot \mathbf{a}$ |
| Associative | $d(\mathbf{a} \cdot \mathbf{b}) = (d\mathbf{a}) \cdot \mathbf{b}$ |
| Distributive | $\mathbf{c} \cdot (\mathbf{a}+\mathbf{b}) = \mathbf{c}\cdot\mathbf{a} + \mathbf{c}\cdot\mathbf{b}$ |
| Zero | $\mathbf{a} \cdot 0 = 0$ |

## The Cross Product

For $\mathbf{a}=[3,2,1]$ and $\mathbf{b}=[1,4,7]$:

1. Stack the vectors
2. Calculate $2 \times 2$ determinants

$$\mathbf{a} \times \mathbf{b} = [2(7)-4(1),\;\; 1(1)-3(7),\;\; 3(4)-2(1)] = [10, -20, 10]$$

- Result is **orthogonal** to both original vectors
- Useful for determinants and matrix inversion

## The Outer Product

Transpose one vector, then multiply:

$$\begin{bmatrix} 3 \\ 2 \\ 1 \end{bmatrix} \begin{bmatrix} 1 & 4 & 7 \end{bmatrix} = \begin{bmatrix} 3 & 12 & 21 \\ 2 & 8 & 14 \\ 1 & 4 & 7 \end{bmatrix}$$

- Input: two vectors of length $k$
- Output: a $k \times k$ matrix

```{r, echo = TRUE}
a <- c(3, 2, 1); b <- c(1, 4, 7)
a %o% b   # outer product
```

## Matrices

A matrix combines row or column vectors. Notation: **bold uppercase** ($\mathbf{A}$).

$$\begin{bmatrix}
a_{11} & a_{12} & \cdots & a_{1n} \\
a_{21} & a_{22} & \cdots & a_{2n} \\
\vdots & \vdots & \ddots & \vdots\\
a_{n1} & a_{n2} & \cdots & a_{nn}
\end{bmatrix}$$

## Matrix Types

| Type | Definition |
|------|-----------|
| **Square** | Equal rows and columns |
| **Symmetric** | Same entries above and below the diagonal; $\mathbf{A} = \mathbf{A}^T$ |
| **Identity** ($\mathbf{I}$) | 1s on diagonal, 0s off; $\mathbf{AI} = \mathbf{A}$ |
| **Idempotent** | $\mathbf{A}^2 = \mathbf{A}$ |
| **Trace** | Sum of diagonal elements: $\text{tr}(\mathbf{I}) = n$ |

## Matrix Addition & Subtraction

Matrices must be **conformable** (same dimensions). Add/subtract element-wise:

$$\begin{bmatrix} a_{11} & a_{12} \\ a_{21} & a_{22} \end{bmatrix} + \begin{bmatrix} b_{11} & b_{12} \\ b_{21} & b_{22} \end{bmatrix} = \begin{bmatrix} a_{11}+b_{11} & a_{12}+b_{12} \\ a_{21}+b_{21} & a_{22}+b_{22} \end{bmatrix}$$

. . .

**Properties:** Commutative, Associative, Distributive, Zero

## Matrix Multiplication
**Order matters!** $\mathbf{AB} \neq \mathbf{BA}$ in general.

Multiply $i$-th row by $j$-th column:

$$\begin{bmatrix} 1 & 3 \\ 2 & 4 \end{bmatrix} \begin{bmatrix} 3 & 5 \\ 2 & 4 \end{bmatrix} = \begin{bmatrix} 1(3)+3(2) & 1(5)+3(4) \\ 2(3)+4(2) & 2(5)+4(4) \end{bmatrix} = \begin{bmatrix} 9 & 17 \\ 14 & 26 \end{bmatrix}$$

```{r, echo = TRUE}
A <- matrix(c(1,3,2,4), nrow=2, byrow=TRUE)
B <- matrix(c(3,5,2,4), nrow=2, byrow=TRUE)
A %*% B
```

. . .

**Conformability rule:** columns of first = rows of second

$$\mathbf{A}_{m \times n} \times \mathbf{B}_{n \times p} = \mathbf{C}_{m \times p}$$

Inner dimensions must match; result has outer dimensions.


## The Transpose

$\mathbf{A}^T$ swaps rows and columns. If $\mathbf{A}$ is $m \times n$, then $\mathbf{A}^T$ is $n \times m$.

$$\begin{bmatrix} 1 & 2 & 3 \\ 4 & 5 & 6 \end{bmatrix}^T = \begin{bmatrix} 1 & 4 \\ 2 & 5 \\ 3 & 6 \end{bmatrix}$$

```{r, echo = TRUE}
A <- matrix(c(1,2,3,4,5,6), nrow=2, byrow=TRUE)
t(A)         # transpose
t(A) %*% A   # A'A is always square & symmetric
```

. . .

**Key properties:**

| Property | Statement |
|----------|-----------|
| Double transpose | $(\mathbf{A}^T)^T = \mathbf{A}$ |
| Sum | $(\mathbf{A}+\mathbf{B})^T = \mathbf{A}^T + \mathbf{B}^T$ |
| Product (**reversal**) | $(\mathbf{AB})^T = \mathbf{B}^T\mathbf{A}^T$ |

## Why the Transpose Matters

**Transposing a product reverses the order:**

$$(\mathbf{ABC})^T = \mathbf{C}^T\mathbf{B}^T\mathbf{A}^T$$

. . .

**Critical result:** For any matrix $\mathbf{A}$, the product $\mathbf{A}^T\mathbf{A}$ is always:

- **Square** ($n \times n$ if $\mathbf{A}$ is $m \times n$)
- **Symmetric**

This is exactly what $\mathbf{X}^T\mathbf{X}$ produces in the normal equations.

## The Determinant

The **determinant** is a **scalar** value computed from a square matrix. It's necessary for *matrix inversion* (later).

**For a $2 \times 2$ matrix:**

$$\det(\mathbf{A}) = \det\begin{bmatrix} a & b \\ c & d \end{bmatrix} = ad - bc$$

. . .

- If $\det(\mathbf{A}) \neq 0$: the matrix is **nonsingular** (invertible)
- If $\det(\mathbf{A}) = 0$: the matrix is **singular** (no inverse exists — columns are linearly dependent)

. . .

**Example:**

$$\det\begin{bmatrix} 4 & 7 \\ 2 & 6 \end{bmatrix} = 4(6) - 7(2) = 10 \neq 0 \;\; ✓$$

$$\det\begin{bmatrix} 2 & 4 \\ 1 & 2 \end{bmatrix} = 2(2) - 4(1) = 0 \;\; \text{(singular — row 1 = 2 × row 2)}$$

```{r, echo = TRUE}
A <- matrix(c(4,7,2,6), nrow=2, byrow=TRUE)
det(A)   # nonsingular
B <- matrix(c(2,4,1,2), nrow=2, byrow=TRUE)
det(B)   # singular — no inverse
```

. . .

**For OLS:** $\det(\mathbf{X}^T\mathbf{X}) = 0$ means perfect multicollinearity — the columns of $\mathbf{X}$ are linearly dependent, and we cannot solve for $\mathbf{b}$.

## Matrix Inversion

For scalars: $a \cdot a^{-1} = 1$

For matrices: $\mathbf{A}\mathbf{A}^{-1} = \mathbf{A}^{-1}\mathbf{A} = \mathbf{I}$

. . .

**Requirements:**

- Only **square** matrices can have inverses
- Must be **nonsingular**: $\det(\mathbf{A}) \neq 0$

## The $2 \times 2$ Inverse

$$\mathbf{A} = \begin{bmatrix} a & b \\ c & d \end{bmatrix} \quad \Longrightarrow \quad \mathbf{A}^{-1} = \frac{1}{ad-bc}\begin{bmatrix} d & -b \\ -c & a \end{bmatrix}$$

. . .

**Example:**

$$\begin{bmatrix} 4 & 7 \\ 2 & 6 \end{bmatrix}^{-1} = \frac{1}{10}\begin{bmatrix} 6 & -7 \\ -2 & 4 \end{bmatrix} = \begin{bmatrix} 0.6 & -0.7 \\ -0.2 & 0.4 \end{bmatrix}$$

$$\mathbf{A}\mathbf{A}^{-1} = \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} = \mathbf{I} \;\; ✓$$

```{r, echo = TRUE}
A <- matrix(c(4,7,2,6), nrow=2, byrow=TRUE)
solve(A)          # inverse
A %*% solve(A)    # verify: should be identity
```

## Properties of the Inverse

| Property | Statement |
|----------|-----------|
| Product (**reversal**) | $(\mathbf{AB})^{-1} = \mathbf{B}^{-1}\mathbf{A}^{-1}$ |
| Transpose | $(\mathbf{A}^T)^{-1} = (\mathbf{A}^{-1})^T$ |
| Double inverse | $(\mathbf{A}^{-1})^{-1} = \mathbf{A}$ |
| Identity | $\mathbf{I}^{-1} = \mathbf{I}$ |

Like the transpose, **inverting a product reverses the order**.

## Linear Regression in Matrix Form

$$\mathbf{y} = \mathbf{X}\mathbf{b} + \mathbf{e}$$

:::: {.columns}

::: {.column width="50%"}

$$\mathbf{y} = \begin{bmatrix} Y_1 \\ Y_2 \\ \vdots \\ Y_n \end{bmatrix} \quad \mathbf{b} = \begin{bmatrix} b_0 \\ b_1 \\ \vdots \\ b_k \end{bmatrix}$$

:::

::: {.column width="50%"}

$$\mathbf{X} = \begin{bmatrix} 1 & X_{11} & \cdots & X_{1k} \\ 1 & X_{21} & \cdots & X_{2k} \\ \vdots & \vdots & \ddots & \vdots \\ 1 & X_{n1} & \cdots & X_{nk} \end{bmatrix}$$

:::

::::

## Deriving the OLS Estimator

**Same Objective:** Minimize the sum of squared errors. In scalar form, we minimized $\sum e_i^2$. In matrix form:

$$\min_{\mathbf{b}}\; \mathbf{e}^T\mathbf{e} = (\mathbf{y} - \mathbf{Xb})^T(\mathbf{y} - \mathbf{Xb})$$

. . .

**Expand** using the distributive property of the transpose, $(\mathbf{A} - \mathbf{B})^T = \mathbf{A}^T - \mathbf{B}^T$:

$$\mathbf{e}^T\mathbf{e} = (\mathbf{y}^T - \mathbf{b}^T\mathbf{X}^T)(\mathbf{y} - \mathbf{Xb})$$

. . .

Multiply out the terms:

$$= \mathbf{y}^T\mathbf{y} - \mathbf{y}^T\mathbf{Xb} - \mathbf{b}^T\mathbf{X}^T\mathbf{y} + \mathbf{b}^T\mathbf{X}^T\mathbf{X}\mathbf{b}$$

. . .

The middle two terms are both scalars (a $1 \times 1$ result), and a scalar equals its own transpose: $\mathbf{y}^T\mathbf{Xb} = (\mathbf{b}^T\mathbf{X}^T\mathbf{y})^T = \mathbf{b}^T\mathbf{X}^T\mathbf{y}$. So they combine:

$$\mathbf{e}^T\mathbf{e} = \mathbf{y}^T\mathbf{y} - 2\mathbf{b}^T\mathbf{X}^T\mathbf{y} + \mathbf{b}^T\mathbf{X}^T\mathbf{X}\mathbf{b}$$

## The Derivative.

In scalar calculus, $\frac{d}{db} f(b)$ gives a single number. In matrix calculus, $\frac{\partial f}{\partial \mathbf{b}}$ gives a **vector of derivatives** — corresponding to each element of $\mathbf{b}$, the intercepts and the slopes

. . .

For instance, if $\mathbf{b} = [b_0, b_1, \ldots, b_k]^T$, then:

$$\frac{\partial f}{\partial \mathbf{b}} = \begin{bmatrix} \frac{\partial f}{\partial b_0} \\ \frac{\partial f}{\partial b_1} \\ \vdots \\ \frac{\partial f}{\partial b_k} \end{bmatrix}$$

This vector of partial derivatives is called the **gradient**. Setting it to $\mathbf{0}$ means *every* partial derivative equals zero simultaneously.

. . .

**Why does this matter?** In the bivariate case, we took two separate derivatives ($\frac{\partial SSR}{\partial a}$ and $\frac{\partial SSR}{\partial b}$) and solved two equations. The gradient is effectively the same thing — but for **all $k+1$ coefficients simultaneously**.

## Matrix Derivative Rules

We need three rules — each corresponds to rules scalar calculus:

. . .

| Scalar Rule | Matrix Rule |
|-------------|-------------|
| $\frac{d}{db}(c) = 0$ | $\frac{\partial}{\partial \mathbf{b}}(\mathbf{y}^T\mathbf{y}) = \mathbf{0}$ |
| $\frac{d}{db}(\mathbf{c}^T b) = \mathbf{c}$ | $\frac{\partial}{\partial \mathbf{b}}(\mathbf{c}^T\mathbf{b}) = \mathbf{c}$ |
| $\frac{d}{db}(b^2 a) = 2ab$ | $\frac{\partial}{\partial \mathbf{b}}(\mathbf{b}^T\mathbf{A}\mathbf{b}) = 2\mathbf{A}\mathbf{b}$ (if $\mathbf{A}$ is symmetric) |

. . .

The third rule requires $\mathbf{A}$ to be symmetric. Since $\mathbf{X}^T\mathbf{X}$ is always symmetric (recall: $(\mathbf{X}^T\mathbf{X})^T = \mathbf{X}^T\mathbf{X}$), the rule applies.

## Applying the Rules

Our function is: $\;\mathbf{e}^T\mathbf{e} = \underbrace{\mathbf{y}^T\mathbf{y}}_{\text{constant}} - \underbrace{2\mathbf{b}^T\mathbf{X}^T\mathbf{y}}_{\text{linear in } \mathbf{b}} + \underbrace{\mathbf{b}^T\mathbf{X}^T\mathbf{X}\mathbf{b}}_{\text{quadratic in } \mathbf{b}}$

. . .

**Term by term:**

| Term | Rule Applied | Derivative w.r.t. $\mathbf{b}$ |
|------|-------------|-------------------------------|
| $\mathbf{y}^T\mathbf{y}$ | Constant → 0 | $\mathbf{0}$ |
| $-2\mathbf{b}^T\mathbf{X}^T\mathbf{y}$ | Linear: $\frac{\partial}{\partial \mathbf{b}}(\mathbf{c}^T\mathbf{b}) = \mathbf{c}$, where $\mathbf{c} = \mathbf{X}^T\mathbf{y}$ | $-2\mathbf{X}^T\mathbf{y}$ |
| $\mathbf{b}^T\mathbf{X}^T\mathbf{X}\mathbf{b}$ | Quadratic: $\frac{\partial}{\partial \mathbf{b}}(\mathbf{b}^T\mathbf{A}\mathbf{b}) = 2\mathbf{A}\mathbf{b}$, where $\mathbf{A} = \mathbf{X}^T\mathbf{X}$ | $2\mathbf{X}^T\mathbf{X}\mathbf{b}$ |

. . .

**Combining:** $\;\frac{\partial\; \mathbf{e}^T\mathbf{e}}{\partial\; \mathbf{b}} = -2\mathbf{X}^T\mathbf{y} + 2\mathbf{X}^T\mathbf{X}\mathbf{b}$

## Setting the Derivative to Zero

Combining the terms:

$$\frac{\partial\; \mathbf{e}^T\mathbf{e}}{\partial\; \mathbf{b}} = -2\mathbf{X}^T\mathbf{y} + 2\mathbf{X}^T\mathbf{X}\mathbf{b} = 0$$

. . .

Rearrange (move $-2\mathbf{X}^T\mathbf{y}$ to the right, divide by 2):

$$\mathbf{X}^T\mathbf{X}\mathbf{b} = \mathbf{X}^T\mathbf{y}$$

These are the **normal equations** — the matrix version of the first-order conditions.

. . .

**Solve for $\mathbf{b}$** by multiplying both sides on the left by $(\mathbf{X}^T\mathbf{X})^{-1}$:

$$(\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{X}\mathbf{b} = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{y}$$

Since $(\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{X} = \mathbf{I}$ and $\mathbf{Ib} = \mathbf{b}$:

$$\boxed{\mathbf{b} = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{y}}$$

This requires $\mathbf{X}^T\mathbf{X}$ to be invertible — fails under **perfect multicollinearity**.

```{r, echo = TRUE}
x <- c(1, 2, 3, 4, 5)
y <- c(5.1, 7.9, 11.2, 13.8, 17.1)
X <- cbind(1, x)
b <- solve(t(X) %*% X) %*% t(X) %*% y
b                    # OLS
coef(lm(y ~ x))     # verify with lm()
```

## Summary

1. Multiple regression extends OLS to multiple predictors: $\hat{Y} = a + b_1 X_1 + b_2 X_2 + \ldots$
2. Matrix algebra provides a compact way to derive the OLS estimator: $\mathbf{b} = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{y}$
3. This requires $\mathbf{X}^T\mathbf{X}$ to be invertible — fails under perfect multicollinearity

## References

- Gill, Jeff. 2006. *Essential Mathematics for Political and Social Research*. Cambridge.
- Moore, Will and David Siegel. 2013. *A Mathematics Course for Political and Social Research*. Princeton.
